---
title: "multivariate models solutions"
author: "Dan McGlinn"
date: "3/24/2016"
output: pdf_document
---

```{r setup, echo=FALSE}
# setup the R enviornment for kniting markdown doc properly
library(knitr)
opts_knit$set(root.dir='../')
```

```{r, tidy=T}
# load code dependancies
library(vegan)
source('./scripts/utility_functions.R')
data(dune)
data(dune.env)
?dune
```

1. Conduct an indirect ordination on the dune plant community. Specifically,
visually examine a NMDS plot using the bray-curtis distance metric. Below is 
some code to help you develop a potential plot that emphasizes the role of the
environmental variable "Moisture". Describe how you interpret the
graphic. What is the goal of creating such a plot? Does this analysis suggest
any interesting findings with respect to the dune vegetation?


```{r, tidy=T}
dune_mds = metaMDS(dune)
# fit enviornmental variables to ordination space
dune_fit = envfit(dune_mds, dune.env)
dune_fit
```

The output from the function `envfit` suggests that "Moisture" and "Management"
are the two most important variables for explaning the site placements in
ordination space. In other words these variables coorelate the strongest to the
primary axes of variation in species composition identified by the analysis. 
Let's examine our oridination plot to see if this seems reasonable.

```{r, tidy=T}
plot(dune_mds, type='n')
text(dune_mds, 'sp', cex=.5)
# generate vector of colors 
mois_lvs = sort(unique(dune.env$Moisture))
color_vect = rev(terrain.colors(length(mois_lvs) + 1))[-1]
points(dune_mds, 'sites', pch=19, 
       col=color_vect[dune.env$Moisture])
# add enviornmental variables for which their p value was less than 0.01
plot(dune_fit, p.max = 0.01, cex=.75)
legend('topright', paste("Moisture =", mois_lvs, sep=''), 
       col=color_vect, pch=19)
```

In the above plot, the points are colored by their Moisture level. It appears
that sites of different moisture levels are arrayed along the first ordination axis (i.e., x-axis) which by definition is the dominant gradient in species composition.
The management levels are arrayed vertically along the second axis indicating that
they are associated with an orthogonal axis of variation in the plant community.

2. Carry out a direct ordination using CCA in order to test any potential 
hypotheses that you developed after examining the MDS plot. Specifically,
carry out a test of the entire model (i.e., including all constrained axes)
and also carry out tests at the scale of individual explanatory variables
you included in your model if you included more than one variable. Plot your 
results.

```{r, tidy=T}

# a very tearse way to specify to use all enviornmental variable is use the "."
# notation, but I don't recommend this because it is not the clearest way to 
# indicate what variables are in the model for a human reader
dune_cca = cca(dune ~ . , data=dune.env)
# alternatively I perfer this specification
dune_cca = cca(dune ~ A1 + Moisture + Management + Use + Manure,
               data=dune.env)
# examine output
dune_cca
```

The output from the CCA model including all the enviornmental variables indicates
the model explains 100 * 1.50 / 2.12 = 71% of the variance in species composition.
However, we included a lot of variables in our model so it is probably a good idea
to compute the adjusted R-squared statistic as well. 

```{r, tidy=T}
dune_cca_r2 = RsquareAdj(dune_cca, nperm=2000)
dune_cca_r2[2]
```

After 3000 permutations you can see that the adjusted r-squared stablized around
`r round(dune_cca_r2$adj.r.squared, 2)` which is quite a bit smaller than the raw r2. This 
indicates that the model was overfit to the data because it had many spurious
explanatory variables. Let's examine if the model and particular variables are statistically significant.

```{r, tidy=T}
# test for model significance
anova(dune_cca)
# test for marginal effects of each variable
anova(dune_cca, by='margin')
```

The first permutation-based ANOVA indicates that the model does explain more 
variance than random, but the effect size is pretty weak as indicated by the small
F-statistic. The second ANOVA which examined the marginal effects of each 
model term indicates that when the variables are forced to compete against one
another that no single variable is statistically significant. The variable "A1"
which is the thickness of the soil A1 horizon is the most important variable 
followed by the Moisture and Manure variables as indicated by their F-statistics.

We could examine to see if single variable models yeild statistically significant
effects but I don't see much point in this excercise. We know that the model 
does capture some of the variance in species composition but that the variables 
in general are fairly weak. We could test the specific hypothesis that was suggested
by the NMDS which was that moisture and management captured orthogonal axes of 
variation in vegetation. 

```{r, tide=T}
dune_cca_MM = cca(dune ~ Moisture +  Management, data=dune.env)
dune_cca_MM_r2 = RsquareAdj(dune_cca_MM, 1000)
dune_cca_MM_r2
anova(dune_cca_MM)
anova(dune_cca_MM, by='margin')
anova(dune_cca_MM, dune_cca)
```

The above analysis indicates that this reduced model model is less overfit 
(note it actually has a higher adjusted r2 because the spurious variables were
dropped). Both moisture and management are relevant variables (i.e., statistically
significant), but their effects are not mind blowing.  

Let's plot our results now:

```{r, tidy=T}
plot(dune_cca)
# bit of a mess, here's a cleaner option to decongest the plot
ordiplot(dune_cca)
# this is cleaner but the blue 'X' are enviornemtnal variables that need labels
# if we want to focus in on the enviornmental variables we can use
plot(dune_cca, display='bp')
```

The second plot indicates that the first axis is primarily being loaded on by 
the linear component of the moisture variable and the cubic component of the
Manure variable. The second axis is more strongly related to differences in 
management as the three classes of management are spread broadly along this axis.

Note that it is difficult to visually compare the strength of a continous 
variable such as A1 with the categorical variables because the relative 
distance of the label seems to be scaled different. The ANOVA analysis is useful
for parsing out the relative importances when different classes of variables
are involved. 

3. Do your two analyses agree with one another or complement one another or do
these two analyses seem to be suggesting different take home messages? Which
analysis do you find to be more useful?

This is a more subjective question but my personal take in this case is that the 
analyses are somewhat complementary. The NMDS is nice because you are ensuring
that you are examining the dominant axes of variation not just the axes that
the enviornment can explain (as in CCA). However, the direct ordination provides
us a clear measures of variable importance and variance explained which are 
intutitive an easier to communicate. With the NMDS bringing in the enviornment 
is always a posthoc approach.  One point to note is that the distance measures 
used in these two analyses are different. In the NMDS analysis the distance 
measure was the bray-curtis distance measure but in the CCA analysis chi-squared
distances are computed. In the case of CCA the chi-squared distances have a 
clear ecological meaning, species are assumed to have an optima along an 
enviornmental gradient and decrease in prevalance away from the optima. In other
words species are expected to have unimodal responses to the enviornment which
is consistent with traditional ecological niche theory. The bray-curtis distance
measure has no such theorectical foundation and its interpretation is thus much
more vauge. 





